Using a Two-Layer Neural Network and Physicochemical Properties to Classify Glass for Forensic Analysis

### 2017-07-02 Paper/paper.tex
Incorporating Mark's Edits

We have chosen a two-layer perceptron network with tangent-sigmoid transfer functions in the hidden layer and \emph{softmax} transfer functions in the output layer. This is a fairly standard network for pattern recognition. Moreover, we will use the \emph{Scaled Conjugate Gradient (SGD)} algorithm to train the network as it is good for pattern recognition problems in which the output layer uses a non-linear transfer function. Since we do not expect the training error to converge to zero, we implement early stopping criteria to prevent overfitting. Lastly, we use \emph{cross-entropy} as our performance index since our targets take on discrete values and it is the optimal performance index for pattern recognition networks that use the \emph{softmax} transfer function in the output layer.

Two different frameworks will be used to implement the neural network. First, we will use the Neural Network Toolbox, specifically the Neural Network Pattern Recognition Tool (\verb|nprtool|) train, validate, and test our network. We use this framework to start with a simple graphical user interface to quickly ensure our specified network architecture is appropriate and to get baseline performance statistics. Then, we will replicate the analysis in Python to gain practical experience building network architectures in a scripting language. Note that since the goal is practical experience, we will not be leveraging the power of the \emph{scikit-learn} package (\verb|sklearn|) for this exercise.

@@

\section{Description of the Network Architecture and Training Algorithm}
\label{sec4}
Once the data is preprocessed, the next step is to decide on or create a network architecture. The basic network architecture is determined by the problem we wish to solve. Once the basic network architecture is determined, we decide how many layers, how many neurons in each layer, how many outputs the network should have, and what kind of performance index function we should use for training \cite{hagan2014}.

\subsection{Network Architecture}
\label{sec4.1}
The standard neural network architecture for pattern recognition problems is the multi-layer perceptron with tangent-sigmoid transfer functions in the hidden layers and \emph{softmax} transfer functions in the output layer. For most problems, including a fairly simple one like ours, one hidden layer usually suffices. Thus, we will implement a two-layer perceptron. If the results of the network are unsatisfactory after training and testing with one hidden layer, we will retrain with an additional hidden layer, but we do not anticipate having to do this. The \emph{tansig} transfer function is usually preferred to the \emph{tansig} transfer function in the hidden layers since it produces outputs (which are inputs to the next layer) that are centered near zero, whereas the \emph{tansig} transfer function always produces positive outputs.

@@

Once we have propagated the input forward, we tcompare the network output to the targets so we can use the error to adjust the weights and biases accordingly. But, in the case of a multi-layer network, the errors and the performance index we use to evaluate those errors are no longer just a function of the weights. Rather, they are indirectly a function of the weights in the hidden layer in addition to the weights in the output layer. Thus, we call the errors we propagate backward through the network `sensitivities' because they represent the sensitivity of the performance index, $F(\mathbf{x})$, to changes in the $i$-th element of the net input at layer $m$. We can calculate these sensitivities analytically as follows: 

@@

&= w_{i, j}^{m+1} \frac{\partial f^{m}(n_{j}^{m}}{\partial n_{j}^{m}} = w_{i, j}^{m+1} \dot{f}^{m}(n_{j}^{m}) ,  

@@

\subsection{Testing the Network}
Testing the network is a simple process since the network is trained and we just need to feed the inputs in to the network. Since the transfer function in the output layer is the \emph{softmax} function, the network output is the class probabilities between 0 and 1. Once the class probabilities are predicted, we need to use these to convert the probabilities to class predictions. To do this, we use 0.5 as the cutoff, so if one neuron receives a higher probability than the other, that one is chosen to be the positive output from the network and the other is assigned to be null. Then, we compare the predictions to the target which tells us the true class. Using this information, we can calculate classification error, accuracy, true and false positive rates, and construct a confusion matrix. We use all of these as performance metrics in discuss them in the next section. The code used to test the network can be found in Appendix F.

@@

This project helped us to understand how flexible and powerful two-layer neural networks can be. Almost any problem can be solved with a two-layer neural network if you have good data, preprocess the data in accordance with the selected network architecture, and choose the right transfer functions and performance index to solve the problem. In class, all of the back propagation problems we solved by hand only had one neuron in the output layer and used a \emph{purelin} transfer function. Building this network in Python without using packages taught us how to use backpropagation with a different performance index and how to derive the sensitivity for the output layer with respect to the output layer and the hidden layer.

>>>>

We have chosen a two-layer perceptron network with \emph{tansig} transfer functions in the hidden layer and \emph{softmax} transfer functions in the output layer. This is a fairly standard network for pattern recognition. Moreover, we will use the \emph{Scaled Conjugate Gradient (SGD)} algorithm to train the network as it is good for pattern recognition problems in which the output layer uses a non-linear transfer function. Since we do not expect the training error to converge to zero, we implement early stopping criteria to prevent overfitting. Lastly, we use \emph{cross-entropy} as our performance index since our targets take on discrete values and it is the optimal performance index for pattern recognition networks that use the \emph{softmax} transfer function in the output layer.

Two different frameworks will be used to implement the neural network. First, we will use the Neural Network Toolbox, specifically the Neural Network Pattern Recognition Tool (\verb|nprtool|) train, validate, and test our network. This framework allows us to quickly ensure our specified network architecture is appropriate and to get baseline performance statistics. Then, we will replicate the analysis in Python to gain practical experience building network architectures in a scripting language. Note that since the goal is practical experience, we will not be leveraging the power of the \emph{scikit-learn} package (\verb|sklearn|) for this exercise.

@@

\section{Description of the Network Architecture and Training Algorithm}
\label{sec4}
Once the data is preprocessed, the next step is to create a network architecture. The basic network architecture is determined by the problem we wish to solve. Once the basic network architecture is determined, we decide how many layers, how many neurons in each layer, how many outputs the network should have, and what kind of performance index function we should use for training \cite{hagan2014}.

\subsection{Network Architecture}
\label{sec4.1}
The standard neural network architecture for pattern recognition problems is the multi-layer perceptron with tangent-sigmoid transfer functions in the hidden layers and \emph{softmax} transfer functions in the output layer. For most problems, including a fairly simple one like ours, one hidden layer usually suffices. Thus, we will implement a two-layer perceptron. If the results of the network are unsatisfactory after training and testing with one hidden layer, we will retrain with an additional hidden layer, but we do not anticipate having to do this. The \emph{tansig} transfer function is usually preferred to the \emph{logsig} transfer function in the hidden layers since it produces outputs (which are inputs to the next layer) that are centered near zero, whereas the \emph{logsig} transfer function always produces positive outputs.

@@

Once we have propagated the input forward, we compare the network output to the targets so we can use the error to adjust the weights and biases accordingly. But, in the case of a multi-layer network, the errors and the performance index we use to evaluate those errors are no longer just a function of the weights. Rather, they are indirectly a function of the weights in the hidden layer in addition to the weights in the output layer. Thus, we call the errors we propagate backward through the network `sensitivities' because they represent the sensitivity of the performance index, $F(\mathbf{x})$, to changes in the $i$-th element of the net input at layer $m$. We can calculate these sensitivities analytically as follows: 

@@

&= w_{i, j}^{m+1} \frac{\partial f^{m}(n_{j}^{m})}{\partial n_{j}^{m}} = w_{i, j}^{m+1} \dot{f}^{m}(n_{j}^{m}) ,  

@@

\subsection{Testing the Network}
Testing the network is a simple process since the network is trained and we just need to feed the inputs in to the network. Since the transfer function in the output layer is the \emph{softmax} function, the network output is the class probabilities between 0 and 1. Once the class probabilities are predicted, we need to use these to convert the probabilities to class predictions. To do this, we use 0.5 as the cutoff, so if one neuron receives a higher probability than the other, that one is chosen to be the positive output from the network and the other is assigned to be zero. Then, we compare the predictions to the target which tells us the true class. Using this information, we can calculate classification error, accuracy, true and false positive rates, and construct a confusion matrix. We use all of these as performance metrics in discuss them in the next section. The code used to test the network can be found in Appendix F.

@@

This project helped us to understand how flexible and powerful two-layer neural networks can be. Almost any problem can be solved with a two-layer neural network if you have good data, preprocess the data in accordance with the selected network architecture, and choose the right transfer functions and performance index to solve the problem. In class, all of the backpropagation problems we solved by hand only had one neuron in the output layer and used a \emph{purelin} transfer function. Building this network in Python without using packages taught us how to use backpropagation with a different performance index and how to derive the sensitivity for the output layer with respect to the output layer and the hidden layer.
